Naturalness and Long-Lived Beyond the Standard Model Particles

### 2017-07-29 tex/draft.tex
Add placeholders for all figures

    In SUSY models, LLPs arise for the same basic reasons, and different SUSY models produce LLPs in different ways. The most straightforward SUSY LLP occurs in models with conserved R-partity, a proposed symmetry that prevents proton decay\fxnote{Be prepared to show this. Maybe add some diagrams illustrating proton decay w/o R-parity}. In models with R-parity, the SM particles are assigned R-parity +1 and their superpartners are assigned R-parity -1. Conserving the product of R-parity at each vertex has two immediate phenomenalogical consequences: (1) sparticles can only be produced in pairs and (2) the lightest supersymmetric particle (LSP) is absolutely stable. A neutral LSP could show up in collider searches as missing energy ($\MET$), making $\MET$ a key signature in many conventional searches.

    \fxnote{Diagrams showing some combination of proton decay, stable LSP, and RPV LSP}

    On the other hand, models with weakkly coupled R-parity violating (RPV) terms produce long-lived but not perfectly stable LSPs. Luckily, RPV models can still avoid proton decay by allowing either terms that violate baryon number or terms that violate lepton number, but not both\fxnote{cite something}. A similar situation arises in GMSB models where the gravitino is the LSP. The coupling between the next-to-LSP (NLSP) and gravitino is inversely proportional to the scale of SUSY breaking, which supresses the NSLP decay rate and can lead to LLPs. \fxnote{is now a good time to explain GMSB?}\fxnote{cite dimopoulos low energy}

@@

\section{The CMS Detector}\fxnote{Add pictures of everything}
\subsection{Overview}
    \fxnote{define eta, coordinate system in general}  
    \fxnote{Mention that ATLAS is similar, says Andrew}  
    \fxnote{Maybe mention PDFs. Definitely know about them, says Andrew}  
    \fxnote{cite cms experiment at the cern lhc}
    The Compact Muon Solenoid (CMS) detector is designed to reconstruct a variety of particles from collisions at the Large Hadron Collider (LHC), which is a circular proton-proton collider that ran with center-of-mass energy \SI{7}{\tera\electronvolt} in 2010-11, \SI{8}{\tera\electronvolt} in 2012, and \SI{13}{\tera\electronvolt} from 2015 to the present. To meet this goal, CMS combines information from several subdetectors nested radially in and around a \SI{4}{\tesla} superconducting solenoid. All together, CMS is \SI{21.6}{\m} long, \SI{14.6}{\m} across, and weighs approximately \SI{12500}{t}. The subdetectors that make up CMS are described below.

\subsection{Tracker}
    \fxnote{cite CMS TDR, jinst 12 c02033}
    \fxnote{mention that analyses below used pre-upgrade tracker}
    In the region closest to the interaction point, CMS uses a fast, high-granularity silicon tracker to reconstruct tracks and pinpoint primary and secondary vertices. The tracker is subdivided into two regions: at $\mathrm{r}<\SI{20}{cm}$, the high particle flux \fxnote{specify} requires small pixels to keep occupancy near \SI{0.1}{\percent}. Moving farther from the IP decreases particle flux, which allows the use of larger silicon strips.

@@

\section{LHC Searches for Long-lived Particles}
    \fxnote{Include diagrams for signal models} 
    \fxnote{mention which data was used for each search} 
    \fxnote{Steal limit plots and any cool figures from papers} 
\subsection{Overview}
    \fxnote{steal jamie's cartoon?}  
    LLPs leave distinct signatures that differentiate them from SM backgrounds but also add complexity to LLP analyses. CMS and ATLAS were mainly designed to reconstruct stable particles from prompt decays, so LLP searches are often forced\fxnote{the searchers are forced, not the searches} to stray from the traditional analysis path in terms of object reconstruction\fxnote{really reconstruction? Maybe object definition?} and background estimation. This report focuses on four complementary LLP signatures: displaced vertices, disappearing tracks, low-velocity particles, and particles that come to a stop before decaying.

@@

    This search\fxnote{repetative wording} uses a simplified benchmark model in the split SUSY paradigm where gluinos are kinematically accessible but squarks are not. The relevant process, shown in \fxnote{reference figure}, is pair-produced gluinos that decay to a quark-\fxnote{how to punctuate?}virtual squark pair. The virtual squark then decays to a quark and the LSP, presumed to be a neutralino\fxnote{have I defined neutralino?}\fxnote{why is this a reasonable assumption?} \fxnote{steal diagram from andrew's preapproval. I can't access it - ask andrew}. The large mass of the intermediate squark suppresses the gluino decay, giving the gluino enough time to form an R-hadron\fxnote{first use. Define here if not in LLP section} that propagates some measureable distance before decaying.

    Jets and MET are reconstructed as in conventional analyses, but dispaced vertices demand special care when identifying tracks. The standard ATLAS tracking algorithm constrains the transverse and longitudinal impact parameters ($\mathrm{d}_0$ and $\mathrm{z}_0$, respectively) of potential tracks, so the authors employ an additional algorithm, referred to as large-radius tracking (LRT). LRT is run after the standard tracking algorithm and uses only hits that were not used in standard track reconstruction. By relaxing requirements on $\mathrm{d}_0$, $\mathrm{z}_0$, and number of hits, LRT is able to reconstruct tracks that point back to DVs.

@@

    The authors minimize the effect of hadronic interactions by vetoing the material-dense regions of the detector\fxnote{cool figure to steal} and estimate the residual background by extrapolating from the low-mass region, which is dominated by hadronic interactions. After a dedicated study, the background from merged vertices was found to be negligable. Finally, the background from crossing tracks is determined with a model that adds pseudo-tracks to data vertices and is normalized by comparing the sample of 3-track vertices in data to (2+1)-track vertices in the model.

    After estimating $\num{0.02} \pm \num{0.02}$ background events, exactly zero are observed. The result is then interpreted as an upper limit on the gluino mass and pair-production cross section as a function of gluino lifetime. For an LSP neutralino mass of \SI{100}{\giga\electronvolt}, the gluino mass limit is approximately \SI{2000}{\giga\electronvolt} for lifetimes in the \SI{20}{\pico\s} to \SI{10}{\nano\s} range. \fxnote{Include plots}

@@

    The authors consider two signal processes: (1) electroweak gaugino production resulting in a disappearing track, an ISR jet, and MET and (2) gluino pair production resulting in a disappearing track, four jets, and MET\fxnote{add diagrams}. Process (1) could be the only gaugino production mode at the LHC if the gluino mass is kinematically unaccessable\fxnote{why?}, but process (2), which leads to chargino production in cascade decays, will be dominant if the gluino mass is within reach.

@@

    No excess over expected background is observed, so upper limits are placed in the NLSP chargino mass-lifetime plane for the EW signal process and in the NLSP chargino mass-gluon mass plane for the QCD production process \fxnote{hopefully will be way clearer after texing in sparticle symbols}. \fxnote{Either include plots or say numbers}

@@

    This analysis is interpreted in the context of two potential LLPs: gluinos and multiply charged massive particles (mchamps). Long-lived gluinos can appear in models such as split\fxnote{capitalize?} SUSY, where the squarks and sleptons are quite heavy, but the gluino is relatively light. After being pair produced, gluinos form R-hadrons\fxnote{have these been defined?} and eventually decay to quark-antiquark pairs along with NLSP neutralinos, which then decay to LSP neutralinos along with muon-antimuon pairs. The gluino decay is supressed by the large squark mass, which leads to its appreciable lifetime.

    This analysis uses the muon system to look for pairs of muons that arrive out of time with bunch crossings. The main backgrounds to such a signal are cosmic muons, beam halo\fxnote{explain}, and muon system noise. Beam halo and muon system noise are easily removed by selection criteria, but distinguishing signal from cosmic muons requires a more nuanced approach. After generating signal MC\fxnote{defined?} and modeling cosmic muons from data obtained in dedicated cosmic runs, the authors found that cosmic muons and signal could be differentiated using TOF information from the muon system RPCs and DTs. Essentially, the direction of muon momentum could be gleaned from the timing information, and the authors required outgoing muons in both the upper and lower hemispheres.\fxnote{Get cartoon from Juliette}

@@

    After estimating the background by extrapolating data from the cosmic-enriched\fxnote{enriched how? be able to explain} region to the signal region and determining systematic uncertainties related to the MC simulation, trigger acceptance, and luminosity, the authors calculate the expected background for LLP lifetimes ranging from \SI{100}{\nano\s} to \SI{1}{\micro\s}. In all cases, the expected background is less then one event. Exactly zero events are observed in every lifetime scenario. These results are then interpreted as upper limits on the signal production cross section\fxnote{and mass limits}\fxnote{include limit plots or table}, which are the first limits for stopped particles that decay to muons at the LHC.

\subsection{Heavy Stable Charged Particles}
    \fxnote{cite CMS EXO7-010}
    \fxnote{Specify dataset}
    \fxnote{nice dE/dx plot (fig 1)}
    Particles that live for more than a couple ns\fxnote{how do I typeset standalone units?} \fxnote{Check this and be more exact} will appear stable in CMS and ATLAS because they can traverse the entire detector before decaying. If such a particle is also sufficiently massive, it can be catigorized as a heavy stable charged particle (HSCP). HSCPs are characterterized by a greater rate of energy loss via ionization due to the lower speed with which they pass through detector material\fxnote{redundant w/ overview. Probably better explained here than there}. Additionally, HSCPs can be fractionally, singly, or multiply charged, which also affects their rate of energy loss.

    HSCPs will often be misidentified or missed entirely by standard reconstruction algorithms and analysis criteria, which tend to assume $\beta \approx \num{1}$ and $\lvert q \rvert = e$. Inverting these assumptions provides two handles for a dedicated HSCP search in the context of three benchmark models: (1) HSCPs that form R-hadrons that either change the sign of their electric charge or become neutral before reaching the MS, (2) lepton-like HSCPs (quasi-stable sleptons) that are produced either through decays of squarks or gluinos or directly if squarks and gluinos are kinematically inaccessible, and (3) long-lived leptonlike fermions with arbitrary electric charge\fxnote{insert feynman diagrams}. This discussion will focus on the first two models, as they bear more directly on naturalness and SUSY\fxnote{why?}.

@@

    No significant excess over predicted background is observed, so the results are interpreted as upper limits in the production cross section - mass plane for the considered models. By comparing the observed limits and theoretical predictions, the authors also extract mass limits for the gluinos, stops, and staus in various models. \fxnote{Include plots}

>>>>

    In SUSY models, LLPs arise for the same basic reasons, and different SUSY models produce LLPs in different ways. The most straightforward SUSY LLP occurs in models with conserved R-partity, a proposed symmetry that prevents proton decay\fxnote{Be prepared to show this. Maybe add some diagrams illustrating proton decay w/o R-parity}. In models with R-parity, the SM particles are assigned R-parity +1 and their superpartners are assigned R-parity -1. Conserving the product of R-parity at each vertex has two immediate phenomenalogical consequences: (1) sparticles can only be produced in pairs and (2) the lightest supersymmetric particle (LSP) is absolutely stable. A neutral LSP could show up in collider searches as missing energy ($\MET$), making $\MET$ a key signature in many conventional SUSY searches.

    On the other hand, models with weakly coupled R-parity violating (RPV) terms produce long-lived but not perfectly stable LSPs. Fig.~\ref{rp_and_rpv_diagrams} shows two possible processes, one where R-parity gaurantees a stable LSP and one where RPV produces a long-lived but stable LSP. A similar situation arises in GMSB models\fxnote{should I describe GMSB?} where the gravitino is the LSP. The coupling between the next-to-LSP (NLSP) and gravitino is inversely proportional to the scale of SUSY breaking, which supresses the NSLP decay rate and can lead to LLPs.\fxnote{cite dimopoulos low energy}\fxnote{}

    \noindent \begin{figure}[htbp] \begin{center}
    \includegraphics[width=0.8\textwidth]{images/rp_and_rpv_diagrams}
        \caption{R-parity and RPV diagrams}
    \label{rp_and_rpv_diagrams}
    \end{center} \end{figure}

@@

\section{The CMS Detector}\fxnote{Add pictures of everything}
    \subsection{Overview}
        \fxnote{define eta, coordinate system in general}  
        \fxnote{Mention that ATLAS is similar, says Andrew}  
        \fxnote{Maybe mention PDFs. Definitely know about them, says Andrew}  
        \fxnote{cite cms experiment at the cern lhc}
        The Compact Muon Solenoid (CMS) detector is designed to reconstruct a variety of particles from collisions at the Large Hadron Collider (LHC), which is a circular proton-proton collider that ran with center-of-mass energy \SI{7}{\tera\electronvolt} in 2010-11, \SI{8}{\tera\electronvolt} in 2012, and \SI{13}{\tera\electronvolt} from 2015 to the present. To meet this goal, CMS combines information from several subdetectors nested radially in and around a \SI{4}{\tesla} superconducting solenoid. All together, CMS is \SI{21.6}{\m} long, \SI{14.6}{\m} across, and weighs approximately \SI{12500}{t}. The subdetectors that make up CMS are shown in Fig.~\ref{cms} and described below.

        \noindent \begin{figure}[htbp] \begin{center}
        \includegraphics[width=0.8\textwidth]{images/cms}
            \caption{CMS}
        \label{cms}
        \end{center} \end{figure}

    \subsection{Tracker}
        \fxnote{cite CMS TDR, jinst 12 c02033}
        \fxnote{mention that analyses below used pre-upgrade tracker}
        In the region closest to the interaction point, CMS uses a fast, high-granularity silicon tracker to reconstruct tracks and pinpoint primary and secondary vertices. The tracker is subdivided into two regions. Inside \SI{20}{cm}, the pixel tracker  uses smaller\fxnote{than what?} silicon pixels to keep the occupancy near \SI{0.1}{\percent} despite the high\fxnote{specify} particle flux. Outside \SI{20}{\cm}, the silicon strip tracker (SST) achieves similar occupancy with larger silicon strips.\fxnote{check occupacy numbers}

@@

\section{LHC Searches for Long-lived Particles}
        \fxnote{Include diagrams for signal models} 
        \fxnote{mention which data was used for each search} 
        \fxnote{Steal limit plots and any cool figures from papers} 
    \subsection{Overview}
        \fxnote{steal jamie's cartoon?}  
        LLPs leave distinct signatures that differentiate them from SM backgrounds but also add complexity to LLP analyses. CMS and ATLAS were mainly designed to reconstruct stable particles from prompt decays, so LLP searches are often forced\fxnote{the searchers are forced, not the searches} to stray from the traditional analysis path in terms of object reconstruction\fxnote{really reconstruction? Maybe object definition?} and background estimation. This report focuses on four complementary LLP signatures: displaced vertices, disappearing tracks, low-velocity particles, and particles that come to a stop before decaying. See Fig.~\ref{jamie_cartoon} for a cartoon of several possible LLP signatures\fxnote{probably a better way to word this sentence}.

        \noindent \begin{figure}[htbp] \begin{center}
        \includegraphics[width=0.8\textwidth]{images/jamie_cartoon}
            \caption{Diagram of several LLP spectra}
        \label{jamie_cartoon}
        \end{center} \end{figure}

@@

        This search\fxnote{repetative wording} uses a simplified benchmark model in the split SUSY paradigm where gluinos are kinematically accessible but squarks are not. The relevant process, shown in Fig.~\ref{displaced_process}, is pair-produced gluinos that decay to a quark-\fxnote{how to punctuate?}virtual squark pair. The virtual squark then decays to a quark and the LSP, presumed to be a neutralino\fxnote{have I defined neutralino?}\fxnote{why is this a reasonable assumption?} \fxnote{steal diagram from andrew's preapproval. I can't access it - ask andrew}. The large mass of the intermediate squark suppresses the gluino decay, giving the gluino enough time to form an R-hadron\fxnote{first use. Define here if not in LLP section} that propagates some measureable distance before decaying.

        \noindent \begin{figure}[htbp] \begin{center}
        \includegraphics[width=0.8\textwidth]{images/displaced_process_diagram}
        \caption{Process.}
        \label{displaced_process}
        \end{center} \end{figure}

        Jets and $\MET$ are reconstructed as in conventional analyses, but dispaced vertices demand special care when identifying tracks. The standard ATLAS tracking algorithm constrains the transverse and longitudinal impact parameters ($\mathrm{d}_0$ and $\mathrm{z}_0$, respectively) of potential tracks, so the authors employ an additional algorithm, referred to as large-radius tracking (LRT). LRT is run after the standard tracking algorithm and uses only hits that were not used in standard track reconstruction. By relaxing requirements on $\mathrm{d}_0$, $\mathrm{z}_0$, and number of hits, LRT is able to reconstruct tracks that point back to DVs.

@@

        The authors minimize the effect of hadronic interactions by vetoing the material-dense regions of the detector (see Fig.~\ref{material_map}) and estimate the residual background by extrapolating from the low-mass region, which is dominated by hadronic interactions. After a dedicated study, the background from merged vertices was found to be negligable. Finally, the background from crossing tracks is determined with a model that adds pseudo-tracks to data vertices and is normalized by comparing the sample of 3-track vertices in data to (2+1)-track vertices in the model.

        \noindent \begin{figure}[htbp] \begin{center}
        \includegraphics[width=0.8\textwidth]{images/material_map}
        \caption{Map of material dense regions of the ATLAS detector.\fxnote{cite analysis}}
        \label{material_map}
        \end{center} \end{figure}

        After estimating $\num{0.02} \pm \num{0.02}$ background events, exactly zero are observed. The result is then interpreted as an upper limit on the gluino mass and pair-production cross section as a function of gluino lifetime, shown in Fig.~\ref{displaced_limit}. For an LSP neutralino mass of \SI{100}{\giga\electronvolt}, the gluino mass limit is approximately \SI{2000}{\giga\electronvolt} for lifetimes in the \SI{20}{\pico\s} to \SI{10}{\nano\s} range. \fxnote{Include plots}

        \noindent \begin{figure}[htbp] \begin{center}
        \includegraphics[width=0.8\textwidth]{images/displaced_limit}
        \caption{Limit.}
        \label{displaced_limit}
        \end{center} \end{figure}

@@

        The authors consider two signal processes, both shown in Fig.\ref{disappearing_processes}: (1) electroweak gaugino production resulting in a disappearing track, an ISR jet, and MET and (2) gluino pair production resulting in a disappearing track, four jets, and MET\fxnote{add diagrams}. Process (1) could be the only gaugino production mode at the LHC if the gluino mass is kinematically unaccessable\fxnote{why?}, but process (2), which leads to chargino production in cascade decays, will be dominant if the gluino mass is within reach.

        \noindent \begin{figure}[htbp] \begin{center}
        \includegraphics[width=0.8\textwidth]{images/disappearing_processes}
        \caption{Processes.}
        \label{disappearing_processes}
        \end{center} \end{figure}

@@

        No excess over expected background is observed, so upper limits are placed in the NLSP chargino mass-lifetime plane for the EW signal process and in the NLSP chargino mass-gluon mass plane for the QCD production process \fxnote{hopefully will be way clearer after texing in sparticle symbols}. Fig.~\ref{disappearing_limits} shows the limits for both processes.

        \noindent \begin{figure}[htbp] \begin{center}
        \includegraphics[width=0.8\textwidth]{images/disappearing_limits}
        \caption{Limits.}
        \label{disappearing_limits}
        \end{center} \end{figure}

@@

        This analysis is interpreted in the context of two potential LLPs: gluinos and multiply charged massive particles (mchamps). Long-lived gluinos can appear in models such as split\fxnote{capitalize?} SUSY, where the squarks and sleptons are quite heavy, but the gluino is relatively light. After being pair produced, gluinos form R-hadrons\fxnote{have these been defined?} and eventually decay to quark-antiquark pairs along with NLSP neutralinos, which then decay to LSP neutralinos along with muon-antimuon pairs, as shown in Fig.~\ref{stopped_process}. The gluino decay is supressed by the large squark mass, which leads to its appreciable lifetime.

        \noindent \begin{figure}[htbp] \begin{center}
        \includegraphics[width=0.8\textwidth]{images/stopped_process}
            \caption{Long-lived gluino proces.\fxnote{cite analysis}}
        \label{stopped_process}
        \end{center} \end{figure}

        This analysis uses the muon system to look for pairs of muons that arrive out of time with bunch crossings. The main backgrounds to such a signal are cosmic muons, beam halo\fxnote{explain}, and muon system noise. Beam halo and muon system noise are easily removed by selection criteria, but distinguishing signal from cosmic muons requires a more nuanced approach. After generating signal MC\fxnote{defined?} and modeling cosmic muons from data obtained in dedicated cosmic runs, the authors found that cosmic muons and signal could be differentiated using TOF information from the muon system RPCs and DTs. Essentially, the direction of muon momentum could be gleaned from the timing information, and the authors required outgoing muons in both the upper and lower hemispheres. Fig.~\ref{stopped_muon_differentiation} shows how muons from stopped particles are differentiated from cosmic muons.

        \noindent \begin{figure}[htbp] \begin{center}
        \includegraphics[width=0.8\textwidth]{images/stopped_muon_differentiation}
            \caption{Cosmic muon rejection with MS.\fxnote{cite analysis}}
        \label{stopped_muon_differentiation}
        \end{center} \end{figure}

@@

        After estimating the background by extrapolating data from the cosmic-enriched\fxnote{enriched how? be able to explain} region to the signal region and determining systematic uncertainties related to the MC simulation, trigger acceptance, and luminosity, the authors calculate the expected background for LLP lifetimes ranging from \SI{100}{\nano\s} to \SI{1}{\micro\s}. In all cases, the expected background is less then one event. Exactly zero events are observed in every lifetime scenario. As shown in Fig.~\ref{stopped_limits}, these results are then interpreted as upper limits on the signal production cross section and mass\fxnote{mass of what?}, which are the first limits for stopped particles that decay to muons at the LHC.

        \noindent \begin{figure}[htbp] \begin{center}
        \includegraphics[width=0.8\textwidth]{images/stopped_limits}
            \caption{Limits.\fxnote{cite analysis}}
        \label{stopped_limits}
        \end{center} \end{figure}

    \subsection{Heavy Stable Charged Particles}
        \fxnote{cite CMS EXO7-010}
        \fxnote{Specify dataset}
        \fxnote{nice dE/dx plot (fig 1)}
        Particles that live for more than a couple ns\fxnote{how do I typeset standalone units?} \fxnote{Check this and be more exact} will appear stable in CMS and ATLAS because they can traverse the entire detector before decaying. If such a particle is also sufficiently massive, it can be catigorized as a heavy stable charged particle (HSCP). HSCPs are characterterized by a greater rate of energy loss via ionization due to the lower speed with which they pass through detector material\fxnote{redundant w/ overview. Probably better explained here than there}. Additionally, HSCPs can be fractionally, singly, or multiply charged, which also affects their rate of energy loss, as shown in Fig.~\ref{hscp_dedx}.

        \noindent \begin{figure}[htbp] \begin{center}
        \includegraphics[width=0.8\textwidth]{images/hscp_dedx}
            \caption{Ionization energy loss.\fxnote{cite analysis}}
        \label{hscp_dedx}
        \end{center} \end{figure}

        HSCPs will often be misidentified or missed entirely by standard reconstruction algorithms and analysis criteria, which tend to assume $\beta \approx \num{1}$ and $\lvert q \rvert = e$. Inverting these assumptions provides two handles for a dedicated HSCP search in the context of three benchmark models\fxnote{do I need figures for these?}: (1) HSCPs that form R-hadrons that either change the sign of their electric charge or become neutral before reaching the MS, (2) lepton-like HSCPs (quasi-stable sleptons) that are produced either through decays of squarks or gluinos or directly if squarks and gluinos are kinematically inaccessible, and (3) long-lived leptonlike fermions with arbitrary electric charge\fxnote{insert feynman diagrams}. This discussion will focus on the first two models, as they bear more directly on naturalness and SUSY\fxnote{why?}.

@@

        No significant excess over predicted background is observed, so the results are interpreted as upper limits in the production cross section - mass plane for the considered models. By comparing the observed limits and theoretical predictions, the authors also extract mass limits for the gluinos, stops, and staus in various models, as shown in Fig.~\ref{hscp_limits}.

        \noindent \begin{figure}[htbp] \begin{center}
        \includegraphics[width=0.8\textwidth]{images/hscp_limits}
            \caption{Limits.\fxnote{cite analysis}}
        \label{hscp_limits}
        \end{center} \end{figure}
